---
title: "Introduction to for loops in R"
author: Will Cornwell & Daniel Falster
output: html_notebook
editor_options: 
  chunk_output_type: console
---

# Introduction

The goal of this prac is to learn to get the computer to repeat things for you and therefore avoid tedium and burnout and stay a happy PhD student with a obedient computer rather than a frustrated, burned-out one with RSI. 

```{r}
for(i in c(3,7,3)){
  print(i)
}
5+3

c(3,7,3)[2]


letters
```



# Your first for-loop

There are datasets in R already.  one of them is called `letters`.  It does this


```{r}
letters

LETTERS
```

write a for loop that prints out each letter sequentially through the alphabet

```{r}
# for (i in c(1,4,5)){
# print(i)
# }

for (i in letters){
print(i)
}
```

extra-credit: write a for loop that prints out every letter except `q`.  never liked `q` anyway.  

```{r}
for(i in letters[letters!="q"]){
  print(i)
}
```

Now look at the directory `data/species`.  It's full of genus level downloads from [GBIF](https://www.gbif.org).   Load all of these csv's into R and combine them into one data frame

```{r}
library(tidyverse)
csv_file_name_vector <- list.files("data/species")
a <- read_csv("data/species/Aonyx.csv")
a

#Putting it in a list

list_of_data <- list()

aa <- read_csv(paste0("data/species/",csv_file_name_vector[1]))

list_of_data[[1]] <- read_csv(paste0("data/species/",csv_file_name_vector[1]))
list_of_data[[2]] <- read_csv(paste0("data/species/",csv_file_name_vector[2]))

dim(list_of_data[[1]])

output_df <- bind_rows(list_of_data)

dim(output_df)

#Ugh this is gonna take forever...what about a loop?

files <- list.files("data/species")

carnivora <- list()

for(i in 1:length(files)){
  carnivora[[i]] <- bind_rows(read_csv(paste0("data/species/", files[i])))
}

carnivora_all <- bind_rows(carnivora)
dim(carnivora_all)
```

Calculate how many species, genera, families, and orders there are in this dataset

```{r}
carnivora_all %>% bind_rows() %>%
  summarise(n_species = length(unique(scientificName)),
            n_genera = length(unique(genus)),
            n_family = length(unique(family)),
            n_order = length(unique(order)))
```


extra-credit: plot the geographic distribution of this dataset, color the points by family

```{r}

```

extra-extra-credit: figure out which species are in Australia and New Zealand for this dataset

```{r}

```
