---
title: "Introduction to for loops in R"
author: "Will Cornwell, Daniel Falster, Fonti Kar, Aniko Toth"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

# Introduction

The goal of this prac is to learn to get the computer to repeat things for you and therefore avoid tedium and burnout and stay a happy PhD student with a obedient computer rather than a frustrated, burned-out one with RSI. 
```{r}
for (i in c(3, 7, 3)) {
  print(i)
}
5 + 3

c(3, 7, 3)[2]


letters
```

# Your first for-loop

There are datasets in R already.  one of them is called `letters`.  It does this

```{r}
letters

LETTERS
```

**Exercises:**
1. Write a for loop that prints out each letter sequentially through the alphabet

```{r, drop}
for (i in letters) {
  print(i)
}
```

2. (stretch) Write a for loop that prints out every letter except `q`.  never liked `q` anyway.  

```{r, drop}
for (i in letters[letters != "q"]) {
  print(i)
}
```

Now look at the directory `data/species`.  It's full of genus level downloads from [GBIF](https://www.gbif.org).   Load all of these csv's into R and combine them into one data frame.

```{r}
library(tidyverse)
csv_file_name_vector <- list.files("data/species")
a <- read_csv("data/species/Aonyx.csv")
a

# Putting it in a list

list_of_data <- list()

aa <- read_csv(paste0("data/species/", csv_file_name_vector[1]))

list_of_data[[1]] <- read_csv(paste0("data/species/", csv_file_name_vector[1]))
list_of_data[[2]] <- read_csv(paste0("data/species/", csv_file_name_vector[2]))

dim(list_of_data[[1]])

output_df <- bind_rows(list_of_data)

dim(output_df)
```

Ugh this is gonna take forever ... what about a loop?

```{r}
files <- list.files("data/species")

carnivora <- list()

for (i in 1:length(files)) {
  carnivora[[i]] <- bind_rows(read_csv(paste0("data/species/", files[i])))
}

carnivora_all <- bind_rows(carnivora)
```

**Exercises:**
1. Calculate how many species, genera, families, and orders there are in this dataset

```{r, drop}
carnivora_all %>%
  summarise(
    n_species = n_distinct(scientificName),
    n_genera = n_distinct(genus),
    n_family = n_distinct(family),
    n_order = n_distinct(order)
  )
```

2. (stretch) plot the geographic distribution of this dataset, color the points by family

```{r, drop}

```

3. (stretch) figure out which species are in Australia and New Zealand for this dataset

```{r, drop}

```
