---
title: "Introduction to `map`"
author: "Will Cornwell, Daniel Falster, Fonti Kar, Aniko Toth"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

# Introduction

The goal of this session is to introduce you to the package [purrr](https://purrr.tidyverse.org) and the `map` family of functions. These provide an alternative and often cleaner way of executing repeated tasks in R than for loops.

To start with, we'll use the same data as the last session.

Load packages
```{r}
library(tidyverse)
```

Here it is, constructed using for loops

```{r}
files <- list.files("data/species", full.names = TRUE)
data <- list()
for (i in seq_along(files)) {
  data[[i]] <- read_csv(files[i])
}
data <- data %>% bind_rows()
```


The alternative way using map is

```{r}
data <- list.files("data/species", full.names = TRUE) %>% map_df(read_csv)
```

Comparing the above two methods, which do you prefer and why?

So we'll now work through how and why the map function works.

# Tidyverse refresh

* The [Tidyverse](http://tidyverse.org) is the name given to suite of R packages designed for seamless data analysis
* Designed to help you fall into a **"Pit of Success"**
* Tools are designed to work seamlessly together, for: 1) Turning data into tidy data, and 2) Plotting & analysing Tidy Data
* Not one but a collection packages
* Dataframes (tibbles) are the universal "tidy" input and output

Load (and install) individually or all together

```{r, eval = FALSE}
library(tidyverse)
```

or 

```{r, eval = FALSE}
library(dplyr)
library(readr)
library(tidyr)
library(ggplot2)
```

Now, let's try counting the number of species in each family

First, try using for loop

```{r}
n <- list()
for (f in unique(data$family)) {
  data_sub <- subset(data, data$family == f)
  n[[f]] <- data.frame(family = f, n = n_distinct(data_sub$scientificName))
}
bind_rows(n)
```

The tidyverse way
```{r}
data %>%
  group_by(family) %>%
  summarise(n = n_distinct(scientificName))
```

## Pipes

The pipe is a bit of magic. It's written by `%>%` (Shift-Command-M on Mac or Shift-Control-M on PC ). We can use "the pipe" [%>%](http://magrittr.tidyverse.org/reference/pipe.html) to connect expressions

```{r}
data$family %>% unique()
```

The pipe 

* `%>%` "pipes" the **output** of the last expression as the **first input** of the next expression
* expects commands on left & right
* If you use RStudio, you can type the pipe with Ctrl + Shift + M if you have a PC or Cmd + Shift + M if you have a Mac.
* you can control the input position of the next function with `.`:

```{r}
2 %>% rnorm(1, 0)
2 %>% rnorm(., 1, 0)

2 %>% rnorm(1, 0, .)

?rnorm() # TIP: To check what arguments you are check
```

**Exercises:** Using the pipe, write expressions to

1. Take the mean of all latitudes  (should be 5.16117)

```{r, drop}
data$decimalLatitude %>% mean()
```

2. List all the species names in alphabetic order

```{r, drop}
data %>%
  arrange(scientificName) %>%
  distinct(scientificName)
```

3. List all the species names in reverse alphabetic order

```{r, drop}
data$scientificName %>%
  unique() %>%
  sort() %>%
  rev()

data %>%
  arrange(scientificName) %>%
  distinct(scientificName)
```


## dplyr workflows

The package `dplyr` is the uses verbs to describe the actions we want to take on the data

- `select`   -> subset columns 
- `filter`   -> subset rows 
- `arrange`  –> order rows
- `rename`   –> rename variables
- `mutate`   –> make new variables
- `summarise`–> summarise data
- `distinct` -> filter to each unique row

Example:

```{r}
data %>%
  select(scientificName, genus, ends_with("tude")) %>%
  filter(genus == "Aonyx") %>%
  summarise(total_n = n())
```


## Developing a dplyr workflow

My strategy for developing a dplyr/pipe workflow

- start with data
- use up key to get last command, add pipe
- keep adding steps
- use right assign if needed
- copy back to source
- reformat if you want
- add assign <- on a different line


Challenge, what to do when doesn't work, which bit?

- to trouble shoot, run the parts

```{r}
data_subset <-
  data %>%
  select(Species, Latitude) %>%
  summarise(
    nspp = n_distinct(Species), 
    max_lat = max(Latitude)
    )
```

**Exercises:** Apply the `dplyr` package and `pipes` to

1. Create a subset of the dataset consisting of only rows for the family "Mustelidae"

```{r, drop}
subset_mustelids <- data %>% filter(family == "Mustelidae")
```

2. As above and only including variables `scientificName`, `decimalLatitude`, `decimalLongitude `

```{r, drop}
subset_mustelids <- 
  data %>%
  filter(family == "Mustelidae") %>%
  select(scientificName, decimalLatitude, decimalLongitude)
```

3. As above but with data sorted  by `decimalLatitude`

```{r, drop}
subset_mustelids <- 
  data %>%
  filter(family == "Mustelidae") %>%
  select(scientificName, ends_with("tude")) %>%
  arrange(decimalLatitude)
```

4. As above but with an additional column  giving the absolute value of `decimalLatitude`

```{r, drop}
subset_mustelids <- 
  data %>%
  filter(family == "Mustelidae") %>%
  select(scientificName, ends_with("tude")) %>%
  arrange(decimalLatitude) %>%
  mutate(abs_decimalLatitude = abs(decimalLatitude))
```

## group_by --  a type of for loop

Another useful dplyr function is `group_by`. This is a type of "for loop"

```{r}
data %>%
  group_by(family) %>%
  summarise(
    n_sp = n_distinct(scientificName),
    n_records = n()
  )
```

How does this work?  (white board)

Split apply combine concept


**Exercises:** 

1. Use dplyr's group_by function to get for each species, the number of records, min and max of decimalLatitude

```{r, drop}
data %>%
  group_by(scientificName) %>%
  summarise(
    n = n(),
    lat_min = min(decimalLatitude),
    lat_max = max(decimalLatitude)
  )
```


# Vectorised functions

Lots of functions naturally work like a for loop. These are what's called vectorised - which means they apply to all elements 

```{r, drop}

data$scientificName %>% str_length()

data$scientificName %>%
  unique() %>%
  str_length()
```

# Purrr: map and walk

Let's now look at the functions included in the package [purrr](https://purrr.tidyverse.org). According to the website, *"purrr enhances R’s functional programming (FP) tool kit by providing a complete and consistent set of tools for working with functions and vectors. If you’ve never heard of FP before, the best place to start is the family of map() functions which allow you to replace many for loops with code that is both more succinct and easier to read."*

Use this when the action you want to repeat can't be achieved in dataframe  piped workflow, or with vectorised function. 

makes use of lists!

loading files
```{r}
files <- list.files("data/species", full.names = TRUE)
data <- list()
for (i in seq_along(files)) {
  data[[i]] <- read_csv(files[i])
}
data <- data %>% bind_rows()
```

Alternative way
```{r}
data <- 
  list.files("data/species", full.names = TRUE) %>%
  map(read_csv) %>%
  bind_rows()
```

How a map routine works works. 3 key steps

- split: break out list or vector into bits
- apply: apply a function to each bit
- combine: combine the results in some fashion (or not)

Each bit needs to be able to run independently 

```{r}
files <- list.files("data/species", full.names = TRUE)

read_csv(files[1])
read_csv(files[2])
read_csv(files[3])
read_csv(files[4])
read_csv(files[5])

map(files, read_csv)

map(files, read_csv) %>% bind_rows()

files %>% map(read_csv)

list.files("data/species", full.names = TRUE) %>%
  map(read_csv) %>%
  bind_rows()
```

A simpler example, print letters 1:10

```{r}
letters[1:10] %>% walk(cat)
```

Important to understand how a list works

**Exercises:** Using the map function

1. Load the first 10 files in the folder "data/species"

```{r, drop}
files <- list.files("data/species", full.names = TRUE)[1:10]

files %>%
  map(read_csv) %>%
  bind_rows()
```


2.  Split the dataset by family -- then count the number of rows. The function `split` is useful here.

```{r, drop}
data %>%
  split(.$family) %>%
  map(nrow)
```


## Map variants

Variants allow for different sorts of return -- i.e. what are you getting back

- map_dbl: returns a number
- map_int: returns an integer
- map_df: returns a dataframe

### Returning an integer

```{r}
data %>%
  split(.$family) %>%
  map(nrow) %>%
  unlist()
```
or 
```{r}
data %>%
  split(.$family) %>%
  map_int(nrow)
```

### Returning a dataframe

```{r}
list.files("data/species", full.names = TRUE) %>%
  map(read_csv) %>%
  bind_rows()
```
or
```{r}
list.files("data/species", full.names = TRUE) %>% map_df(read_csv)
```

### map2 -- when you have 2 inputs

Let's copy files to another directory, use the function `file.copy`

Here we copy one file.
```{r}
dir.create("output")

file.copy("data/species/Aonyx.csv", "output/data-Aonyx.csv")
```

Now let's copy all of them by calling map:

```{r}
# 
location1 <- list.files("data/species", full.names = TRUE)
location2 <- 
  location1 %>% str_replace_all("data/species/", "output/data-")

walk2(location1, location2, file.copy)
```

