---
title: "Writing your own functions in R"
author: "Will Cornwell, Daniel Falster, Fonti Kar, Aniko Toth"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

# Introduction

The goal of this session is to learn how to write your own functions in R. 

Three big reasons to write functions

1. Using function to reuse code / avoid repeated code
2. Using function to tidy up your analysis
3. Using (anonymous) functions to extend possibilities in dplyr or map workflows 

Research is by definition new, and so sometimes you have to do things that no one has done before.  If you do that new thing more than once, it's worth writing it into a function

Use similar packages and data to yesterday 
```{r}
library(tidyverse)
```

Introduce new dataset - gapminder

```{r}
data_gapminder <- read_csv("data/gapminder/gapminder-FiveYearData.csv")
```

## First, let's inspect the dataset

Challenges:

- run some basic summaries on the gapminder data
  -  how many countries, years, continents are included in the data?
  - what is the range of `lifeExp` & `gdpPercap` for each country?
  - can you plot `lifeExp` vs `gdpPercap` in year 1992?
  - can you plot `lifeExp` vs `gdpPercap` in year 1992, faceted by continent?  

```{r, drop}

```


## Now some examples on what we can achieve

Example 1 - tidy & turbo charge model fitting

```{r}
fit_model <- function(data) {
  lm(lifeExp ~ gdpPercap, data = data)
}

get_r2 <- function(fit) {
  summary(fit)$r.squared
}

fits <-
  data_gapminder %>%
  split(.$country) %>%
  map(fit_model)

map_dbl(fits, get_r2)
```


Example 2 - tidy & turbo charge plotting

```{r}
make_plot_year <- function(this_year, data) {
  data %>%
    filter(year == this_year) %>%
    ggplot(aes(gdpPercap, lifeExp, colour = continent, size = log10(pop))) +
    geom_point(alpha = 0.5) +
    scale_x_log10(limits = c(200, 200000)) +
    scale_y_log10(limits = c(20, 100)) +
    theme_minimal() +
    labs(x = "GDP per capita", y = "Life expectancy (yrs)", title = this_year)
}

make_plot_year(1982, data_gapminder) %>% print()
```

The printing is kind of annoying, so could make another function:

```{r}
make_plot_year_print <- function(...) {
  make_plot_year(...) %>% print()
}

make_plot_year_print(1982, data_gapminder)
```

No let's use our new skills to make lots of plots!

```{r}
years <- unique(data_gapminder$year)

walk(years, make_plot_year_print, data_gapminder)

for (year in years) {
  make_plot_year_print(year, data_gapminder)
}
```

Or we could save the plots
```{r}
save_plot <- function(this_year, data_gapminder, path = "output/by_year/") {
  dir.create(path, FALSE, TRUE)
  make_plot_year(this_year, data_gapminder)
  ggsave(paste0(path, this_year, ".png"))
}

for (year in years) {
  save_plot(year, data_gapminder)
}

years %>% walk(save_plot, data_gapminder)
```

# Structure of a function

function structure (rowSums)

- name
- body
- arguments

investigating structure of function f 

- `args(f)`
- `body(f)` or just `f`
- return value
- ?help
- named vs ordered arguments
- default arguments

Here's a simple example, implementing a simple function

```{r}
mymean <- function(x) {
  sum(x) / length(x)
}
```

**Exercises:**  

Load carnivore data
```{r}
carnivore_data <- list.files("data/species", full.names = TRUE) %>% map_df(read_csv)
```

1. Extract the most northern observations for each species in the southern hemisphere Write a function that finds the largest value that is below 0

```{r, drop}
find_largest_negative <- function(x) {
  max(x[x < 0])
}

carnivore_data %>% 
  group_by(scientificName) %>%
  summarise(most_northern_south = find_largest_negative(decimalLatitude))
```

2. If your function returns -Inf for some cases, change what it returns to NA possibly using `na_if` or `ifelse` inside the function.  This is called handling edge cases ellegantly.  Of course this can be done outside the function, but it's a better, more robust function if this is done internally.  

```{r, drop}
find_largest_negative_better <- function(x) {
  y <- max(x[x < 0])
  na_if(y, -Inf)
}

# OR

find_largest_negative_better <- function(x) {
  if (all(x > 0)) {
    y <- NA
  } else {
    y <- max(x[x < 0])
  }
  return(y)
}


carnivore_data %>% 
  group_by(scientificName) %>%
  summarise(
    north_south = find_largest_negative_better(decimalLatitude)
    )
```


# Combining functions with map

From yesterday, we had this example

```{r}
list.files("data/species", full.names = TRUE) %>% map_df(read_csv)
```

What if we want to record which files each data point came from? Write a function that loads a file and adds the filename in a new column of the data

```{r, drop}
load_file <- function(x) {
  read_csv(x) %>% mutate(filename = x)
}

carnivore_data <- list.files("data/species", full.names = TRUE) %>% map_df(load_file)
```

**Exercises:**  We'll now make some plots

some pretty code to use 

```{r}
world_map <- map_data("world")
ggplot() +
  geom_polygon(
    data = world_map, aes(x = long, y = lat, group = group),
    fill = "darkgrey", colour = "white"
  )
```

1. Write a function that takes the dataframe and a family name and plots the distribution for each family and saves it to the output file in a folder called "figures"

```{r, drop}
dir.create("figures")

plot_family <- function(df, family) {
  df_specific_family <-
    filter(df, family == family)
  world_map <- map_data("world")
  ggplot(df_specific_family, aes(x = decimalLongitude, y = decimalLatitude)) +
    geom_polygon(
      data = world_map, aes(x = long, y = lat, group = group),
      fill = "darkgrey", colour = "white"
    ) +
    geom_point() +
    ggtitle(family)
  ggsave(paste0("figures/", family, ".pdf"))
}

plot_family(carnivore_data, "Odobenidae")
```

2. Now re-write this function so it works with map, and make a plot of all the families

```{r, drop}
plot_families <- function(fam, df) {
  df_specific_family <- filter(df, family == fam)
  world_map <- map_data("world")
  ggplot(df_specific_family, aes(x = decimalLongitude, y = decimalLatitude)) +
    geom_polygon(
      data = world_map, aes(x = long, y = lat, group = group),
      fill = "darkgrey", colour = "white"
    ) +
    geom_point() +
    ggtitle(fam)
  ggsave(paste0("figures/", fam, ".pdf"))
}

family_list <- as.list(unique(carnivore_data$family))

map(family_list, plot_families, df = carnivore_data)
```

3. Some of the observations in GBIF are only identified to genus. Figure out how many this is and report for this dataset the percentage.  `str_to_sentence` or `toupper` or `word` might be useful.

```{r, drop}
carnivore_data$binom <- word(carnivore_data$scientificName, 1, 2)
sum(carnivore_data$binom != str_to_sentence(carnivore_data$binom), na.rm = T) / length(carnivore_data$binom)

problems <- filter(carnivore_data, binom != str_to_sentence(binom))
```


4. Your colleague asks you to match your data against theirs. They have used 6-letter abbreviations for species, i.e.. "urvjav" for "Urva javanica". Write a function to separate out the first two words from scientificName and use a map function to apply this to the full species list

```{r, drop}
species <- carnivore_data$scientificName %>%
  unique() %>%
  sort()

very_short_name <- function(x) {
  paste0(
    word(x, 1, 1) %>% str_sub(1, 3),
    word(x, 2, 2) %>% str_sub(1, 3)
  )
}

carnivore_data$scientificName %>% map_chr(very_short_name)
```

## Anonymous functions

Remember we did this:

```{r}
load_file <- function(x) {
  read_csv(x) %>% mutate(filename = x)
}

list.files("data/species", full.names = TRUE) %>% map_df(load_file)
```


Another way:

```{r}
list.files("data/species", full.names = TRUE) %>%
  map_df(function(x) read_csv(x) %>% mutate(filename = x))


# Or
list.files("data/species", full.names = TRUE) %>%
  map_df(~ read_csv(.x) %>% mutate(filename = .x))
```


**Exercises:** Reimplement the following example as an anonymous function

```{r, drop}
short_name <- function(x) {
  x2 <- x %>% str_split(" ")
  paste(x2[[1]][1], x2[[1]][2])
}

carnivore_data$scientificName %>% map_chr(short_name)
```


## Writing longer functions

Store these in another file and load using `source`

Identify the bit that will differ and 


**Exercises:** Write a function called `load_data` that loads the carnivora data and creates a column with short name and very short name

```{r, drop}

load_file <- function(x) {
  read_csv(x,
    col_types =
      cols(
        scientificName = col_character(),
        family = col_character(),
        decimalLatitude = col_double(),
        decimalLongitude = col_double(),
        genus = col_character(),
        order = col_character()
      )
  ) %>%
    mutate(filename = x)
}


load_data <- function() {
  list.files("data/species", full.names = TRUE) %>%
    map_df(load_file) %>%
    mutate(
      species = scientificName %>% map_chr(short_name),
      species6 = scientificName %>% map_chr(very_short_name)
    )
}

data <- load_data()
```

## Extra exercises

1. Write your own function to convert `decimalLatitude` to degrees minutes and seconds e.g 42.3601 becomes 42° 21' 36.36"
  - For the degrees use the whole number part of the decimal
  - For the minutes multiply the remaining decimal by 60. Use the whole number part of the answer as minutes.
  - For the seconds multiply the new remaining decimal by 60
  - Output degrees, minutes and seconds as separate columns

```{r, drop}
carnivore_data$decimalLatitude[[1]]
str_detect(as.character(carnivore_data$decimalLatitude[[1]]), ".")

convert_DMS <- function(data, var) {
  # Split up first numeric string by decimal place
  parts <- str_split(as.character(carnivore_data[[var]]), "\\.", 2)

  # Assign the first element as degrees
  degrees <- parts[[1]][1] %>% as.numeric()

  # Add back in the decimal place and multiple second element by 60 and assign as minutes
  minutes <- paste0(".", parts[[1]][2]) %>% as.numeric() * 60

  # Split minutes by decimal place for seconds
  second_part <- minutes %>%
    as.character() %>%
    str_split("\\.", 2)

  # Multiple values after the decimal place by 60 and assign as second
  seconds <- paste0(".", second_part[[1]][2]) %>% as.numeric() * 60

  # The return
  data %>% mutate(
    degrees = degrees,
    minutes = minutes,
    seconds = seconds
  )
}

convert_DMS(carnivore_data, decimalLatitude)
```
